---
title: "Antibiotics"
author: "Satchel"
date: 2020-
output:
  github_document
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Creating effective data visualizations is an *iterative*
process; very rarely will the first graph you make be the most
effective. The most effective thing you can do to be successful in this
iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely
dependent on *the question you are trying to answer*. A visual that is
totally ineffective for one question may be perfect for answering a
different question.

In this challenge, you will practice *iterating* on data visualization,
and will anchor the *assessment* of your visuals using two different
questions.

*Note*: Please complete your initial visual design **alone**. Work on
both of your graphs alone, and save a version to your repo *before*
coming together with your team. This way you can all bring a diversity
of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the
[*minimum inhibitory
concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration)
(MIC) of three drugs for different bacteria. The smaller the MIC for a
given drug and bacteria pair, the more practical the drug is for
treating that particular bacteria. An MIC value of *at most* 0.1 is
considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin,
streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized
into a genus based on a number of features, including their resistance
to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

```{r}
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,

    plot.margin = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    plot.title = element_text(size = 18),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 12)
  )
}
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the
data** `df_antibiotics`. These **cannot** be simple variations on the
same graph; for instance, if two of your visuals could be made identical
by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16
bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of
ideas, and don't fixate on any one idea for too long. You will have a
chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be
helpful to `pivot_longer()` the data to make certain visuals easier to
construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE
df_antibiotics_longer <-pivot_longer(df_antibiotics, cols = c(penicillin, streptomycin, neomycin), 
                                     names_to = "antibiotic", 
                                     values_to = "effectiveness")
df_antibiotics_longer

df_antibiotics_longer %>% 
  ggplot(aes(x = bacteria, y = effectiveness, shape = antibiotic, color = gram)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(17, 18, 15)) +
  scale_color_manual(values = c("red", "blue"), labels = c("Gram-positive", "Gram-negative")) +
  labs(x = "Bacteria", y = "MIC Score", shape = "Antibiotic", color = "Gram Status") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE
df_antibiotics_longer %>% 
ggplot( aes(x = bacteria, y = effectiveness, fill = antibiotic)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~gram, scales = "free_y") +
  labs(title = "Antibiotic effectiveness by bacteria and gram stain",
       x = "Bacteria", y = "MIC Score",
       fill = "Antibiotic") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE
df_antibiotics_longer %>% 
ggplot( aes(x = antibiotic, y = effectiveness)) + 
  geom_boxplot(fill = "#69b3a2") +
  facet_wrap(~bacteria, scales = "free_y") +
  labs(title = "Effectiveness of Antibiotics for Different Bacteria",
       x = "Antibiotic",
       y = "Effectiveness")
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE

df_antibiotics_longer %>% 
  ggplot( aes(x = antibiotic, y = bacteria, fill = effectiveness)) + 
  geom_tile() + 
  scale_fill_gradient(low = "white", high = "black") +
  labs(title = "Antibiotic Effectiveness", x = "Antibiotic", y = "Bacteria")

df_antibiotics_longer %>% 
ggplot(aes(x = antibiotic, y = bacteria, color = antibiotic)) +
  geom_point(aes(size = effectiveness)) +
  scale_color_discrete(name = "antibiotic") +
  labs(x = "antibiotic", y = "bacteria") +
  theme_classic()
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE


df_antibiotics_longer %>% 
  ggplot(aes(x = antibiotic, y = effectiveness, fill = gram)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#1B9E77", "#D95F02")) +
  labs(x = "Antibiotic", y = "Effectiveness", fill = "Gram Stain") +
  ggtitle("Effectiveness of Antibiotics Against Bacteria") +
  theme_bw() +
  geom_point(aes(x = antibiotic, y = effectiveness, color = bacteria), size = 2, alpha = 0.7, stat = "identity")

```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer
both Guiding Questions. Note that you must also identify which of your
five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in
answering the questions below. You may need to revise one or more of
your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most
effective at helping answer both guiding questions. **Use this as an
opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against
> bacteria of different genera and Gram stain?

*Observations* - What is your response to the question above? - (Write
your response here) - Which of your visuals above (1 through 5) is
**most effective** at helping to answer this question? - (Write your
response here) - Why? - (Write your response here)

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*,
and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*
[2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus
> pneumoniae*?

*Observations* - What is your response to the question above? - (Write
your response here) - Which of your visuals above (1 through 5) is
**most effective** at helping to answer this question? - (Write your
response here) - Why? - (Write your response here)

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide
antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009)
[link](https://www.americanscientist.org/article/thats-funny)
